Statistically-estimated tree composition for the northeastern United States \\ at the time of European settlement

### 2015-08-20 paper.tex
added final edits from coauthors and cleaned up some formatting

\& These authors contributed equally to this work. 
\end{quote}
TODO:

focus on PLOS One, but Nature's Scientific Data is a possibility

check journal requirements for archiving and presentation; talk to
Jody/Ann about archiving/DOI

\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the northeastern United States
and the statistical methodology used to produce the product. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height in 22 taxonomic groupings, generally
at the genus level. The data come from settlement survey records that
provide raw data that are transcribed and then aggregated spatially,
giving count data. The domain is divided into two regions, eastern
(Maine to Ohio) and western (Indian to Minnesota). Public Land Survey
point data in the western region are aggregated to a regular 8 km
grid, while data in the eastern region, from Town Proprietor Surveys,
is aggregated at the township level in irregularly-shaped local administrative
units. The product is based on a Bayesian statistical model fit to
the count data that estimates composition on a regular 8 km grid.
The statistical model allows us to estimate composition at locations
with no data and to smooth over noise caused by limited counts in
locations with data. Critically, it also allows us to quantify uncertainty
in our composition estimates. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

Historical datasets provide critical context to understand forest
ecology. They allow researchers to define 'baseline' conditions for
conservation management, to understand ecosystem processes at decadal
and centennial scales, and, particularly in regions with widespread
land use change, to understand the extent to which forests after conversion
and regeneration differ from the original forest cover.

Euro-American settlement and subsequent land use change occured in
a time transient fashion across North America, and land surveys that
provide vegetation information followed a similar pattern. Early surveys
(from 1620 until 1825) in the northeastern United States provide areally-aggregated
data at the township level \citep{cogbill2002forests,thompson2013four};
we refer to these as the Town Proprietor Survey (TPS). These surveys
provide composition of major tree species using sometimes inconsistent
common names. Later surveys, after the establishment of the U.S. Public
Land Survey System (PLS) - from 1832 to 1907 - provide point-level
data along a regular grid, with one mile spacing \citet{bourdo1956review,schulte2001original,goring2015composition},
again with sometimes inconsistent common names. Survey instructions
during the PLS varied through time and by point type. This requires
the application of spatially-varying correction factors \citep{cogbill2015corrections,goring2015composition}
to accurately assess stem density, basal area and biomass from the
early settlement records but has limit impact when estimating composition.

Logging, agriculture, and abandonment have left an indelible mark
on forests in the northeastern United States \citet{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \citep{rhemtulla2009historical,friedman2005regional}
and with varying spatial resolution, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a dataset for the upper Midwest that is resolved to an 8 x
8 km grid size, providing broad spatial coverage at a spatial scale
that can be compared to the Forest Inventory and Analysis products
\citet{gray2012forest}. Combined with additional, coarsely-sampled
PLS data from Illinois and Indiana and with the TPS data, this gives
us raw data for much of the northeastern United States. However, there
are several limitations of using the raw data that can be alleviated
by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data provide only estimates of
within-cell variance that do not account for information from nearby
locations. Second, the available digitized data from Illinois and
Indiana represent a small fraction of those states and missing townships
are common in the TPS data.

Properly assessing uncertainty in ecological data is imperative to
understanding and modelling ecological processes \citep{cressie2009accounting}.
In this way, a model that can account for the spatial structure of
the underlying PLS and TPS data, and provide reliable estimates of
uncertainty across the northeastern United States, provides a valuable
tool for researchers interested in the ecological structure and function
of forests at longer time scales.

\section{Data}

The raw data were obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at PLS survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. (Note
that for Indiana and Illinois, at the moment trees are associated
with township centroids and then assigned to 8 km grid cells based
on the centroid but in the near future we will have point locations
available for each tree.) For the states of Ohio, Pennsylvania, New
Jersey, New York and the six New England states (the eastern subdomain),
data are aggregated at the township level. There are also data from
a single township in Quebec and a single township in northern Delaware.
Data are essentially complete in Minnesota, Wisconsin and Michigan,
but data in Illinois and Indiana represent a sample of the full set
of grid cells, with survey record transcription ongoing. Data for
the remaining states are available for a subset of the full set of
townships covering the domain. Fig. \ref{fig:domain} {[}Fig. 1 -
Lyx is having a numbering problem{]} shows the domain, indicating
the grid cells and townships with data.

\begin{figure}
\label{fig:domain}\includegraphics{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}

\caption{Spatial domain, with locations with data shown in gray. Locations
are grid cells in western portion and townships in eastern portion.
In addition to locations without data being indicated in white, grid
cells completely covered in water are white (e.g., a few locations
in Minnesota and Wisconsin).}
\end{figure}

@@

Extensive details on the data are available in \cite{Gori:etal:2015}
and the raw data are available at XXX {[}ask Simon/Charlie about public
access to Paleon composition data{]}. The aggregation into taxonomic
groups is primarily at the genus level, but is at the species level
in some cases of monospecific genera. We model the following 22 taxa
plus an ``other hardwood'' category: Atlantic white cedar (\emph{Chamaecyparis
thyoides}), Ash (\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}),
Beech (\emph{Fagus grandifolia)}, Birch (\emph{Betula spp.}), Black
gum/sweet gum (\emph{Nyssa sylvatica} and \emph{Liquidambar styraciflua}),
Cedar/juniper (\emph{Juniperus virginiana} and \emph{Thuja occidentalis}),
Cherry (\emph{Prunus spp.}), Chestnut (\emph{Castanea dentata}), Dogwood
(\emph{Cornus spp.}), Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}),
Hemlock (\emph{Tsuga canadensis}), Hickory (\emph{Carya spp.}), Ironwood
(\emph{Carpinus caroliniana} and \emph{Ostrya virginiana}), Maple
(\emph{Acer spp.}), Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}),
Poplar/tulip poplar (\emph{Populus spp.} and \emph{Liriodendron tulipifera}),
Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), Walnut
(\emph{Juglans nigra}). Note that in several cases (e.g., black gum/sweet
gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different families and even orders. For the western subdomain
we do not fit statistical models for Atlantic white cedar and chestnut
as these have 0 and 6 trees present, respectively. The taxa grouped
into the other hardwood category are those for which fewer than roughly
2000 trees were present in the dataset; however we include Atlantic
white cedar explicitly despite it only having 336 trees in the dataset.
Given the nature of the data, there is not a firm cutoff on the sizes
of trees included, but the data roughly represent trees greater than
8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
but with some trees as small as 4 inches. {[}Charlie,Simon, please
check this wording{]} In Indiana and Illinois there are also 402 trees
and 29 trees, respectively, that are between 1 and 4 inches dbh, but
these will be excluded in the next version of the product.

@@

Our domain is a rectangle covering all of the states, in the Albers
projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
rectangle split into 8 km cells, arranged in a 296 by 180 grid of
cells, with the centroid of the cell in the southeast corner located
at (-71000, 58000). For the modeling of the western subdomain we use
the western-most 146 by 180 grid of cells. For the modeling of the
eastern subdomain we use the eastern-most 180 by 180 grid of cells
and then omit 23 cells in the north and 17 cells in the south outside
the states containing data.

\section{Statistical model}

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using full population of all the trees in
an areal region.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
Data Product section), and can then be used in subsequent analyses.
The mean and standard deviation of the samples for each cell by taxon
pair represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

@@

We start by describing the basic model for those states for which
we have raw data on the 8 km grid and in Section \ref{sub:Model-for-township}
we describe the extension of the model to accommodate data aggregated
at the township level.

The statistical model treats the observations as coming from a multinomial
distribution with a (latent) vector of proportions for each grid cell.
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i}))
\]
where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
is the vector of unknown proportions for those taxa at that cell.
Note that we use a standard multinomial distribution without overdispersion
as the set of trees in the dataset is roughly uniformly sampled across
the cells or townships \citep{Gori:etal:2015}.

@@

MRF models work directly with the precision matrix of the values of
the spatial process, so calculation of the prior density of $\alpha_{p}(s)$
is computationally simple \citep{Rue:Held:2005}, but in situations
where the likelihood is not normal, it can be difficult to set up
effective MCMC algorithms that are able to move in the high-dimensional
space of $\alpha_{p}$. The latent variable representation of Section
\ref{sub:Model-for-township} helps to alleviate this problem.

\subsubsection{Standard conditional autoregressive models}

Our first model is a standard conditional autoregressive (CAR) model
\citep{Bane:etal:2003}. We use a standard form of this model, which
treats the four cardinal neighbors of each grid cell as the neighbors
of the grid cell. The corresponding precision matrix has diagonal
elements, $Q_{ii}$, equal to the number of neighbors for the $i$th
area (i.e., four except for cells on the boundary of the domain),
while $Q_{ik}=-1$ (the negative of a weight of one) when areas $i$
and $k$ are neighbors and $Q_{ik}=0$ when they are not. This gives
the following model for the values of $\alpha_{p}(s_{i})$ collected
as a vector across all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-})
\]
The use of the generalized inverse notation indicates that $Q$ is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and we can write $Q=D-C$ where $C$ is the $m\times m$ adjacency
matrix defining the neighborhood relation of the locations; that is
$(C)_{ik}=1$ if locations $i$ and $k$ are neighbors. The matrix
$D$ is an $m\times m$ diagonal matrix containing the row sums of
matrix $C$ as the diagonal entries $(D)_{ii}={\displaystyle \sum_{k=1}^{m}(C)_{ik}}.$

@@

\subsubsection{Gaussian process approximation}

@@

The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
to the Matern-based GP for $\nu=1$ and $\nu=2$. Our second model
is the Lindgren approximation for Matern-based GPs with $\nu=1$.
To implement the Lindgren model, one modifies the $Q$ matrix defined
previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

@@

\noindent The ICAR specification contains a set of hyperparameters
$\sigma_{p}^{2}$ for $p$ = $1,...,P$. Following \cite{Gelm:2006}
we use a uniform distribution on each $\sigma_{p}$ parameter, with
upper bound of {[}check max values{]}. For the SPDE model we also
have parameters $\mu_{p}$, which we give flat, non-informative priors
(truncated at $\pm10$), and $\rho_{p}$ which we give uniform priors
on the interval $(0.1,\exp(5))$.

@@

It is well-known that devising an effective MCMC algorithm for models
with latent Gaussian process(es) and a non-Gaussian likelihood is
difficult (cite recent cent/noncent papers \citep{Rue:Held:2005,Chri:etal:2006,Tan:Nott:2013}.
To develop an algorithm, we make use of a latent variable representation
for the multinomial probit model \citep{McCu:Ross:1994}. The representation
introduces latent variables that allow one to develop a MCMC sampling
strategy that takes advantage of closed form full conditional distributions
(so-called Gibbs sampling steps) for $\alpha_{p}$.

Suppose that compositional counts are available at a number of locations.
At location $i$, a sample size of $n_{i}$ observations are collected,
and each observation (i.e., each tree) can be classified into $P$
distinct categories. For a given tree $j$ at location $i$, let $Y_{ij}$
denote the response variable indicating the category. Let $Y_{ij}$
be associated with $P$ latent variables $W_{ij1},...,W_{ijP}$ such
that $Y_{ij}$ = $p$ if and only if $W_{ijp}={\displaystyle \max_{p'}\big\{ W_{ijp'}\big\}}$;
in other words, the maximum of the set of latent variables $\{W_{ijp}\}{\displaystyle _{p=1}^{P}}$
determines the category of observation $j$ at location $i$.

\noindent The final piece of the latent variable representation is
the relationship between the $W$ variables and the $\alpha_{p}(s)$
processes. We have that

\noindent 
\[
W_{ijp}\sim\mbox{N}(\alpha_{p}(s_{i}),1)
\]
independently for all of the $W_{ijp}$ values.

\noindent Consider the following example with two locations that are
neighbors and $P=2$ categories. Each tree $j$ at location $i$ is
associated with two variables $W_{ij1}$ and $W_{ij2}$, governed
by the latent variables $\alpha_{1i}$ and $\alpha_{2i}$, respectively.
Suppose that $\alpha_{1i}>\alpha_{2i}$ for a given location $i$.
Then this model implies that any tree $j$ is more likely to be labeled
1 than 2 at location $i$. The difference between $\alpha_{1i}$ and
$\alpha_{2i}$ explains the \textit{difference} in probability of
\textit{categories} 1 and 2 at location $i$, and the similarity between
$\alpha_{p1}$ and $\alpha_{p2}$ explains the \textit{correlation}
between the probabilities at \textit{locations} 1 and 2 for category
$p$.

@@

We developed an extension of the model described in previous sections
to account for data at a different aggregation than our core 8 km
grid. This extension introduces a new set of latent variables, one
per tree, that indicate the grid cell in which the tree is located.
These latent 'membership' variables, $c_{tj}$, for tree $j$ in township
$t$, $t=1,\ldots,T$ can be sampled within the MCMC as additional
unknown parameters. The prior for $c_{tj}$ is a discrete distribution
that puts mass proportional to the areal overlap between the township
in which the tree is located and the $m$ grid cells, with the priors
across different trees in a township being independent and with most
of the . 
\[
c_{tj}\sim\mbox{Multinom}(1,\{\psi_{1},\ldots,\psi_{m}\}).
\]
Because the townships overlap a limited number of grid cells, most
of the $\psi_{1},\ldots,\psi_{m}$ values are zero.

@@

While the latent variable representation provides great advantages
in the MCMC sampling for each $\alpha_{p}$ compared to joint Metropolis
updates or updating each location individually, there is still strong
dependence between the hyperparameters, $\{\sigma_{p}^{2},\mu_{p},\rho_{p}\}$
and the latent process values (as well as between the latent process
values and the latent $W_{ijp}$ variables). To address the first,
we developed a 'cross-level' joint updating strategy in which we propose
$\phi\in\{\sigma_{p}\},p=1,\ldots,P$ (and for the SPDE model, $\phi\in\{\{\sigma_{p}\},\{\mu_{p}\},\{\rho_{p}\}\}),$
via a Metropolis-style random walk and then conditional on the proposed
value of $\phi^{*}$ propose $\alpha_{p}$ from its full conditional
distribution given $\phi^{*}$and the latent $W$ variables. This
is equivalent to sampling from the marginalized (with respect to $\alpha_{p}$)
distribution of $\phi$ conditional on $W_{p}$. Note that in sampling
$\sigma_{p}$and $\rho_{p}$ in the SPDE model, for each $p$, we
jointly propose $\{\sigma_{p},\rho_{p}\}$ using an adaptive Metropolis
proposal and then use the cross-level strategy to also propose $\alpha_{p}$
as part of a single accept-reject step. For these various joint samples
of hyperparameters and $\alpha_{p}$, we use adaptive Metropolis sampling
\citep{Shab:Well:2011}.

@@

The model is implemented in R \cite{R:2014} with core computational
calculations coded in C++ using the \emph{Rcpp} package \citep{Edde:Fran:2011}.
We also make extensive use of sparse matrix representations and algorithms,
using the \emph{spam} package in R \citep{Furr:Sain:2010}. All code
is available on Github, including pre- and post-processing code, at
https://github.com/Paleon-Project/composition.

>>>>

\& These authors contributed equally to this work.

Short title: Settlement-era northeast U.S. tree composition
\end{quote}

\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the northeastern United States
and the statistical methodology used to produce the product. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height in 22 taxonomic groupings, generally
at the genus level. The data come from settlement survey records that
provide raw data that are transcribed and then aggregated spatially,
giving count data. The domain is divided into two regions, eastern
(Maine to Ohio) and western (Indiana to Minnesota). Public Land Survey
point data in the western region are aggregated to a regular 8 km
grid, while data in the eastern region, from Town Proprietor Surveys,
are aggregated at the township level in irregularly-shaped local administrative
units. The product is based on a Bayesian statistical model fit to
the count data that estimates composition on a regular 8 km grid.
The statistical model allows us to estimate composition at locations
with no data and to smooth over noise caused by limited counts in
locations with data. Critically, it also allows us to quantify uncertainty
in our composition estimates. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

Historical datasets provide critical context to understand forest
ecology. They allow researchers to define `baseline' conditions for
conservation management, to understand ecosystem processes at decadal
and centennial scales, and, particularly in regions with widespread
land use change, to understand the extent to which forests after conversion
and regeneration differ from the original forest cover.

Euro-American settlement and subsequent land use change occurred in
a time-transient fashion across North America, and land surveys that
provide vegetation information followed a similar pattern. Early surveys
(from 1620 until 1825) in the northeastern United States provide areally-aggregated
data at the township level \citep{cogbill2002forests,thompson2013four};
we refer to these as the Town Proprietor Survey (TPS). Later surveys,
after the establishment of the U.S. Public Land Survey System (PLS)
provide point-level data along a regular grid, with one mile spacing
during the period 1832 to 1907 \citep{bourdo1956review,schulte2001original,goring2015composition}.
Survey instructions during the PLS varied through time and by point
type. Accounting for this requires the application of spatially-varying
correction factors \citep{cogbill2015corrections,goring2015composition}
to accurately assess stem density, basal area and biomass from the
early settlement records, but the impact on composition estimates
is limited. Both types of surveys provide composition of major tree
species but using sometimes inconsistent common names.

Logging, agriculture, and land abandonment have left an indelible
mark on forests in the northeastern United States \citep{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \citep{friedman2005regional,rhemtulla2009historical}
and with varying spatial resolution, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a dataset for the upper Midwest that is resolved to an 8 x
8 km grid size, providing broad spatial coverage at a spatial scale
that can be compared to the modern forests using Forest Inventory
and Analysis products \citep{gray2012forest}. Combined with additional,
coarsely-sampled PLS data from Illinois and Indiana and with the TPS
data, this gives us raw data for much of the northeastern United States.
However, there are several limitations of using the raw data that
can be alleviated by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data provide only estimates of
within-cell variance that do not account for information from nearby
locations. Second, the available digitized data from Illinois and
Indiana represent a small fraction of those states and missing townships
are common in the TPS data. Our statistical model allows us to provide
a spatially-complete data product of settlement-era tree composition
with uncertainty across the northeastern U.S.

In Section \ref{sec:Data} we describe the data sources, while Section
\ref{sec:Statistical-model} describes our statistical models. In
Section \ref{sec:Model-comparison} we quantitatively compare competing
statistical specifications, and in Section \ref{sec:Data-product}
we describe the final data product. In Section \ref{sec:Discussion}
we discuss limitations of the statistical model, and we list related
data products under development.

\section{Data\label{sec:Data}}

The raw data were obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at PLS survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. (Note
that for Indiana and Illinois, at the moment trees are associated
with township centroids and then assigned to 8 km grid cells based
on the centroid but in the near future we will have point locations
available for each tree.) For the states of Ohio, Pennsylvania, New
Jersey, New York and the six New England states (the eastern subdomain),
data are aggregated at the township level. There are also data from
a single township in Quebec and a single township in northern Delaware.
Data are essentially complete in Minnesota, Wisconsin and Michigan,
but data in Illinois and Indiana represent a sample of the full set
of grid cells, with survey record transcription ongoing. Data for
the remaining states are available for a subset of the full set of
townships covering the domain. Fig. \ref{fig:Spatial-domain} shows
the domain, indicating the grid cells and townships with data.

\begin{figure}
\label{fig:domain}\includegraphics[scale=1.3]{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}

\caption{Spatial domain of the northeastern United States, with locations with
data shown in gray. Locations are grid cells in western portion and
townships in eastern portion. In addition to locations without data
being indicated in white, grid cells completely covered in water are
white (e.g., a few locations in the northwestern portion of the domain
in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
\end{figure}

@@

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available in \cite{Gori:etal:2015}.
The lower Midwest (Illinois, Indiana) data were purchased from the
Indiana State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. The TPS data were compiled by C.V. Cogbill from a myriad of
archival sources representing land division surveys conducted in connection
with local settlement. The complete raw vegetation data are available
through the NIS Data Portal (accession number and version to be inserted
when available). The aggregation into taxonomic groups is primarily
at the genus level, but is at the species level in some cases of monospecific
genera. We model the following 22 taxa plus an ``other hardwood''
category: Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash
(\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus
grandifolia)}, Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa
sylvatica} and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
virginiana} and \emph{Thuja occidentalis}), Cherry (\emph{Prunus spp.}),
Chestnut (\emph{Castanea dentata}), Dogwood (\emph{Cornus spp.}),
Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}), Hemlock (\emph{Tsuga
canadensis}), Hickory (\emph{Carya spp.}), Ironwood (\emph{Carpinus
caroliniana} and \emph{Ostrya virginiana}), Maple (\emph{Acer spp.}),
Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}), Poplar/tulip
poplar (\emph{Populus spp.}~and \emph{Liriodendron tulipifera}),
Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), and
Walnut (\emph{Juglans nigra}). Note that in several cases (black gum/sweet
gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different families and even orders. For the western subdomain
we do not fit statistical models for Atlantic white cedar and chestnut
as these have 0 and 7 trees present, respectively. The taxa grouped
into the other hardwood category are those for which fewer than roughly
2000 trees were present in the dataset; however we include Atlantic
white cedar explicitly despite it only having 336 trees in the dataset.
Although surveyors avoided using small trees, there was no consistent
lower diameter limit. The data, however, roughly represent trees greater
than 8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
but with some trees as small as 3 inches dbh. In Indiana and Illinois
there are also 402 trees and 29 trees, respectively, that are between
1 and 4 inches dbh, but these will be excluded in the next version
of the product {[}waiting on Simon to see if he stripped them out{]}.

@@

Our domain is a rectangle covering all of the states, in the Albers
projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
rectangle split into 8 km cells, arranged in a 296 by 180 grid of
cells, with the centroid of the cell in the southwest corner located
at (-71000, 58000). For the modeling of the western subdomain we use
the western-most 146 by 180 grid of cells. For the modeling of the
eastern subdomain we use the eastern-most 180 by 180 grid of cells
and then omit 23 cells in the north and 17 cells in the south outside
the states containing data.

\section{Statistical model\label{sec:Statistical-model}}

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using full population of all the trees in
an areal region.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
Data Product section) and can then be used in subsequent analyses.
The mean and standard deviation of the samples for each pair of cell
and taxon represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

@@

We start by describing the basic model for those states for which
we have raw data on the 8 km grid, and in Section \ref{sub:Model-for-township}
we describe the extension of the model to accommodate data aggregated
at the township level.

The statistical model treats the observations as coming from a multinomial
distribution with a (latent) vector of proportions for each grid cell,
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i})),
\]
where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
is the vector of unknown proportions for those taxa at that cell.
Note that we use a standard multinomial distribution without overdispersion
as the set of trees in the dataset is roughly uniformly sampled across
the cells or townships \citep{Gori:etal:2015}.

@@

MRF models work directly with the precision matrix of the values of
the spatial process, so calculation of the prior density of $\alpha_{p}$
is computationally simple \citep{Rue:Held:2005}, but in situations
where the likelihood is not normal, it can be difficult to set up
effective MCMC algorithms that are able to move in the high-dimensional
space of $\alpha_{p}$. The latent variable representation of Section
\ref{sub:Model-for-township} helps to alleviate this problem.

\paragraph{Standard conditional autoregressive models}

Our first model is a standard conditional autoregressive (CAR) model
\citep{Bane:etal:2003}. We use a standard form of this model, which
treats the four cardinal neighbors of each grid cell as the neighbors
of the grid cell. The corresponding precision matrix has diagonal
elements, $Q_{ii}$, equal to the number of neighbors for the $i$th
area (i.e., four except for cells on the boundary of the domain),
while $Q_{ik}=-1$ (the negative of a weight of one) when areas $i$
and $k$ are neighbors and $Q_{ik}=0$ when they are not. This gives
the following model for the values of $\alpha_{p}(s_{i})$ collected
as a vector across all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-}).
\]
The use of the generalized inverse notation indicates that $Q$ is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and we can write $Q=D-C$ where $C$ is the $m\times m$ adjacency
matrix defining the neighborhood relation of the locations; that is
$C{}_{ik}=1$ if locations $i$ and $k$ are neighbors. The matrix
$D$ is an $m\times m$ diagonal matrix containing the row sums of
matrix $C$ as the diagonal entries $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$

@@

\paragraph{Gaussian process approximation}

@@

The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second model
is the Lindgren approximation for \matern-based GPs with $\nu=1$.
To implement the Lindgren model, one modifies the $Q$ matrix defined
previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

@@

\noindent The ICAR specification contains a set of hyperparameters
$\{\sigma_{p}^{2}\},$ $p=1,\ldots P$. Following \cite{Gelm:2006}
we use a uniform distribution on each $\sigma_{p}$ parameter, with
upper bound of 1000. For the SPDE model we also have parameters $\mu_{p}$,
which we give flat, non-informative priors (truncated at $\pm10$),
and $\rho_{p}$ which we give uniform priors on the interval $(0.1,\exp(5))$.

@@

It is well-known that devising an effective MCMC algorithm for models
with latent Gaussian process(es) and a non-Gaussian likelihood is
difficult \citep{Rue:Held:2005,Chri:etal:2006,Tan:Nott:2013}. To
develop an algorithm, we make use of a latent variable representation
for the multinomial probit model \citep{McCu:Ross:1994}. The representation
introduces latent variables that allow one to develop a MCMC sampling
strategy that takes advantage of closed form full conditional distributions
(so-called Gibbs sampling steps) for $\alpha_{p}$.

Suppose that compositional counts are available at a number of locations.
At location $i$, a sample of size $n_{i}$ observations is collected,
and each observation (i.e., each tree) can be classified into $P$
distinct categories. For a given tree $j$ at location $i$, let $Y_{ij}$
denote the response variable indicating the category. Let $Y_{ij}$
be associated with $P$ latent variables $W_{ij1},...,W_{ijP}$ such
that $Y_{ij}$ = $p$ if and only if $W_{ijp}={\displaystyle \max_{p'}\big\{ W_{ijp'}\big\}}$;
in other words, the maximum of the set of latent variables $\{W_{ijp}\}{\displaystyle _{p=1}^{P}}$
determines the category of observation $j$ at location $i$. The
final piece of the latent variable representation is the relationship
between the $W$ variables and the $\alpha_{p}(s)$ processes. We
have that

\noindent 
\[
W_{ijp}\sim\mbox{N}(\alpha_{p}(s_{i}),1)
\]
independently for all of the $W_{ijp}$ values. Consider the following
example with two locations that are neighbors and $P=2$ categories.
Each tree $j$ at location $i$ is associated with two variables $W_{ij1}$
and $W_{ij2}$, governed by the latent variables $\alpha_{1i}$ and
$\alpha_{2i}$, respectively. Suppose that $\alpha_{1i}>\alpha_{2i}$
for a given location $i$. Then this model implies that any tree $j$
is more likely to be labeled 1 than 2 at location $i$. The difference
between $\alpha_{1i}$ and $\alpha_{2i}$ explains the \textit{difference}
in probability of \textit{categories} 1 and 2 at location $i$, and
the similarity between $\alpha_{p1}$ and $\alpha_{p2}$ explains
the \textit{correlation} between the probabilities at \textit{locations}
1 and 2 for category $p$.

@@

We developed an extension of the model described in previous sections
to account for data at a different aggregation than our core 8 km
grid. This extension introduces a new set of latent variables, one
per tree, that indicate the grid cell in which the tree is located
and can be sampled within the MCMC as additional unknown parameters.
Specifically, $c_{tj}$, is the latent `membership' variable for tree
$j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$ is
a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
proportional to the areal overlap between the township in which the
tree is located and the $m$ grid cells, giving 
\[
c_{tj}\sim\mbox{Multinom}(1,\{\psi_{t1},\ldots,\psi_{tm}\}),
\]
independently across all trees. Because the townships overlap a limited
number of grid cells, most of the $\psi_{t1},\ldots,\psi_{tm}$ values
are zero.

@@

While the latent variable representation provides great advantages
in the MCMC sampling for each $\alpha_{p}$ compared to joint Metropolis
updates or updating each location individually, there is still strong
dependence between the hyperparameters, $\{\sigma_{p}^{2},\mu_{p},\rho_{p}\}$
and the latent process values (as well as between the latent process
values and the latent $W_{ijp}$ variables). To address the first,
we developed a `cross-level' joint updating strategy for the CAR model
in which we propose $\phi_{p}=\sigma_{p},p=1,\ldots,P,$ (and for
the SPDE model, $\phi_{p}\in\{\{\sigma_{p},\rho_{p}\},\mu_{p}\}$)
via a Metropolis-style random walk and then conditional on the proposed
value of $\phi_{p}^{*}$ propose $\alpha_{p}$ from its full conditional
distribution given $\phi_{p}^{*}$ and the latent $W_{p}$ variables.
This is equivalent to sampling from the marginalized (with respect
to $\alpha_{p}$) distribution of $\phi_{p}$ conditional on $W_{p}$.
For these various joint samples of hyperparameters and $\alpha_{p}$,
we use adaptive Metropolis sampling \citep{Shab:Well:2011}.

@@

The model is implemented in R \citep{R:2014} with core computational
calculations coded in C++ using the \emph{Rcpp} package \citep{Edde:Fran:2011}.
We also make extensive use of sparse matrix representations and algorithms,
using the \emph{spam} package in R \citep{Furr:Sain:2010}. All code
is available on Github, including pre- and post-processing code, at
https://github.com/Paleon-Project/composition.
